### 2020-06-03 report/sections/server/codegen/implementation.tex
Report: progress on chapter 4

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
Its rich standard library also simplifies many tasks: we use
the \texttt{argparse} package to generate an informative
command line interface (CLI) for developers to 
supply the correct information (\cref{lst:cmdline})
to use \fancyname{SessionTS},
and \texttt{subprocess} package provides an abstraction to
invoke external tools, such as Scribble and the TypeScript Compiler.

>>>>

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
Its rich standard library also simplifies many tasks: we use
the \texttt{argparse} package to generate an informative
command line interface (CLI) for developers to 
supply the correct information (\cref{lst:cmdline})
to use \fancyname{SessionTS},
and the \texttt{subprocess} package to 
invoke external tools, such as Scribble and the TypeScript Compiler.

### 2020-06-03 report/sections/server/codegen/workflow.tex
Report: progress on chapter 4

\begin{figure}[!ht]
\begin{lstlisting}[language=Scribble]
module Adder;

global protocol Adder(role Client, role Svr) {
	choice at Client {
		ADD(number, number) from Client to Svr;
		RES(number)         from Svr to Client;	
	} or {
		QUIT(string) from Client to Svr;	
		TERMINATE()  from Svr to Client;
	}
}
\end{lstlisting}
\captionof{lstlisting}{The \tprotocol{Adder} Protocol}
\label{lst:adder}
\end{figure}

We will use the \tprotocol{Adder} protocol as a running example
to demonstrate how our work performs TypeScript API generation.

@@

As the code generation process requires additional information,
we define an \texttt{Endpoint} dataclass\footnote{
A Python dataclass uses the decorator to generate
``boilerplate'' methods, such as the constructor, based on the
properties listed in the annotations.} (\cref{lst:endpointapi})
to contain the \texttt{EFSM}
representation, along with the information passed in from the
command line (\texttt{protocol}, \texttt{role}, \texttt{server}) 
and the custom type declarations (\texttt{types}) parsed from the
protocol specification.

>>>>

\begin{figure}[!ht]
\begin{lstlisting}[language=Scribble]
global protocol Adder(role Client, role Svr) {
	choice at Client {
		ADD(number, number) from Client to Svr;
		RES(number)         from Svr to Client;
		do Adder(Client, Svr);
	} or {
		QUIT(string) from Client to Svr;
		choice at Svr {
			THANKS()    from Svr to Client;
		} or {
			TERMINATE() from Svr to Client;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{The \tprotocol{Adder} Protocol}
\label{lst:adder}
\end{figure}

We will use the \tprotocol{Adder} protocol as a running example
to demonstrate how our work performs TypeScript API generation.
This describes a binary session between \trole{Client} and \trole{Svr}:
if \trole{Client} asks \trole{Svr} to \tmsg{ADD} two numbers,
\trole{Svr} responds with the \tmsg{RES}ult; this loop continues
until \trole{Client} kindly asks to \tmsg{QUIT} with a message,
where \trole{Svr} either responds kindly with \tmsg{THANKS},
or just says \tmsg{TERMINATE}, but both choices will close the session.

@@

As the code generation process requires additional information,
we define an \texttt{Endpoint} dataclass\footnote{
A Python dataclass uses the \texttt{@dataclass} decorator to generate
``boilerplate'' methods, such as the constructor, based on the
properties listed in the annotations.} (\cref{lst:endpointapi})
to contain the \texttt{EFSM}
representation, along with the information passed in from the
command line (\texttt{protocol}, \texttt{role}, \texttt{server}) 
and the custom type declarations (\texttt{types}) parsed from the
protocol specification.

### 2020-06-03 report/sections/server/node.tex
Report: progress on chapter 4

\begin{figure}[!b]
\centering
%\includegraphics[width=0.5\textwidth]{•}
\captionof{figure}{\dots}
\label{fig:addersvrfsm}
\end{figure}

@@

We motivate our approach from \cite{PureScript2019, HuJava}
to generate \textbf{handler-style APIs} to be implemented
by developers.
This aligns with idiomatic TypeScript practices of defining
\textit{callbacks} in the application logic.
By strictly specifying handlers for send and receive actions,
we do not expose send and receive APIs to the developer,
making it impossible for the developer to reuse channels.

@@

When executing \fancyname{SessionTS} to generate code for the
\trole{Svr} endpoint specifying the \texttt{node} target,
we obtain the following files:

\begin{itemize}
\item \textbf{EFSM.ts:} \dots
\item \textbf{Svr.ts:} \dots
\end{itemize}

\section{State Encodings}
\label{section:nodeefsm}

@@

\subsection{Send}
\begin{itemize}
\item very short -- go over the structure and give examples
\item show how it extends for polyadic payloads
\end{itemize}

We generalise deterministic send actions as a trivial \textit{selection}, 
as motivated from the theory (\dots).

\subsection{Receive}
\begin{itemize}
\item very short -- go over the structure and give examples
\item show how it extends for polyadic payloads
\end{itemize}

>>>>

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{AdderSvrFSM}
\captionof{figure}{\trole{Svr} Endpoint FSM
in \tprotocol{Adder} protocol}
\label{fig:addersvrfsm}
\end{figure}

@@

We motivate our approach from \cite{PureScript2019, HuJava}
to generate \textbf{handler-style APIs} to be implemented
by developers.
This aligns with idiomatic TypeScript practices of defining
\textit{callbacks} in the application logic.
We type the parameters and return values of the handlers to
reflect the message types specified in the protocol.
By strictly specifying handlers for send and receive actions,
we do not expose send and receive APIs to the developer,
making it impossible for the developer to reuse channels.

The responsibility of guaranteeing linear channel usage now falls
under the runtime that executes the EFSM.
As we generate this for the developer,
we can provide \textbf{static}
linearity guarantees by construction.

@@

When executing \fancyname{SessionTS} to generate code for the
\trole{Svr} endpoint specifying the \texttt{node} target,
the developer obtains the following files:

\begin{itemize}
\item \textbf{EFSM.ts:} TypeScript types constructed for 
the EFSM encoding (\cref{section:nodeefsm});
\item \textbf{Svr.ts:} Session runtime for executing the EFSM 
(\cref{section:noderuntime}).
\end{itemize}

\section{EFSM Encoding}
\label{section:nodeefsm}

In addition to states and transitions, we generate 
TypeScript constructs
for relevant metadata in the state machine: this includes
\textit{roles}, \textit{labels} and \textit{message structures}.

The EFSM contains more than just states and transitions

The generated \textbf{EFSM.ts} file encodes all information
available

\begin{figure}
\begin{lstlisting}[language=javascript,tabsize=2]
export namespace Roles {...};
export namespace Labels {...};
export namespace Message {...};

export namespace Handler {...};

abstract class ISend {...};
abstract class IReceive {...};
abstract class ITerminal {...};
export namespace Implementation {...};

export type EfsmTransitionHandler =
	(implementation: Implementation.Type) => void;
export type MessageHandler = (message: any) => void;
\end{lstlisting}
\captionof{lstlisting}{Structure of generated EFSM.ts file 
for server endpoints}
\label{lst:nodeefsmfile}
\end{figure}

@@

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, tabsize=2]
// Inside the Message namespace...
export type S54THANKS = {
	label: Labels.S54.THANKS,
	payload: [string],
};
export type S54TERMINATE = {
	label: Labels.S54.TERMINATE,
	payload: [],
};

export type S54 = | S54THANKS | S54TERMINATE;
\end{lstlisting}
%\label{}
\captionof{lstlisting}{Generated Message Type Definition for State 54}
\end{figure}

\subsection{Handler APIs}

We collect the APIs that the developer needs to implement
under the \texttt{Handler} namespace. 
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.
We outline how the generated API differs between sending
and receiving states.

\paragraph{Send}
We generalise deterministic send actions as a trivial \textit{selection}, 
as motivated from the theory (\dots).

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, tabsize=2]
// Inside the Handler namespace...
export type S54 = 
	| [Labels.S54.THANKS, Message.S54THANKS['payload'],
		Implementation.S52] 
	| [Labels.S54.TERMINATE, Message.S54TERMINATE['payload'], 
		Implementation.S52];
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Send State
in \tprotocol{Adder} protocol}
\end{figure}

\paragraph{Receive}
\begin{itemize}
\item very short -- go over the structure and give examples
\item show how it extends for polyadic payloads
\end{itemize}

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside the Handler namespace...
export type S51 = {
	[Labels.S51.ADD]: (...payload: Message.S51ADD['payload']) =>
		Implementation.S53,
	[Labels.S51.QUIT]: (...payload: Message.S51QUIT['payload']) => 
		Implementation.S54,
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Svr} Receive State
in \tprotocol{Adder} protocol}
\end{figure}
